Comparison of several sequential and parallel derivative-free global optimization algorithms

### 2018-11-26 2019/PAVT/pavt_paper.tex
PAVT:  add unrelated references, fix table size

В данной работе рассмотрены open-source реализации девяти различных методов глобальной оптимизации.
Все алгоритмы были протестированы на наборе из 900 существенно многоэкстремальных функций, который был сгенерирован с
помощью специализированных генераторов задач \cite{Gaviano2003, grishaginClass}. Помимо сравнения последовательных алгоритмов, на подмножестве из 200 тестовых задач произведено сравнение солвера MIDACO \cite{Schlueter2012} и программной системы Globalizer \cite{globalizerSystem} в условиях работы на суперкомпьютере Лобачевский.
\end{Russian}

@@

Using the multiple mapping allows solving initial problem (\ref{eq:task}) by parallel solving the
problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1\leqslant s\leqslant S
\]
on a set of intervals $[0,1]$ by the index method. Each one-dimensional problem is solved on a
separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by
particular processor are interpreted as the results of the trials in the rest problems (in the
corresponding points \(x^{k_1},\dots,x^{k_S})\). In this approach, a trial at the point \(x^k \in
[0,1]\) executed in the framework of the \(s\)-th problem, consists in the following sequence of
operations:
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point \( y^k\) (the
blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1\leqslant s\leqslant S\), of the point \(y^k\) and interpret the
trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points
\(x{}^{k_1} ,\dots,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par
The decision rules for the mentioned parallel algorithm, in general, are the same as the rules of the
sequential algorithm (except the method of the trial execution). Each processor has its own copy
of the software realizing the computations of the problem functions and the decision rule of the
index algorithm. For the organization of the interactions among the processors, the queues are
created on each processor, where the processors store the information on the executed iterations
in the form of the tuples: the processor number \(s\), the trial point \(x{}^{k_s}\).
\par
The mentioned parallelization scheme was proposed in \cite{Gergel2009} and implemented in the Globalizer system with the use of MPI technology. Main
features of implementation consist in the following:
\begin{itemize}
  \item A separate MPI-process is created for each
 of \(S\) one-dimensional problems being solved, usually, one process per one processor
 employed.
 \item Each process can use $p$ threads to parallel execute $p$ trials, usually one thread per an accessible core.
\end{itemize}

@@

\begin{table}
\begin{center}
\caption{Trials limit for the each of test problems classes}
  \begin{tabular}{|l|{c}|}
    \hline
  Class & Limit\\
  \hline
  \(F_{GR}\) & 5000 \\
  \hline
  GKLS 2d Simple & 8000 \\
  \hline
  GKLS 2d Hard & 9000 \\
  \hline
  GKLS 3d Simple & 15000 \\
  \hline
  GKLS 3d Hard & 25000 \\
  \hline
  GKLS 4d Simple & 150000 \\
  \hline
  GKLS 4d Hard & 250000 \\
  \hline
  GKLS 5d Simple & 350000 \\
  \hline
  GKLS 5d Hard & 600000 \\
  \hline
  \end{tabular}
  \label{tab:limits}
\end{center}
\end{table}

В качестве характеристик метода оптимизации на каждом из классов будем рассматривать среднее число
испытаний, затраченное для решения одной задачи, и количество решённых задач. Чем меньше число испытаний, тем быстрее метод сходится
к решению, а значит и меньше обращается к потенциально трудоёмкой процедуре вычислений целевой функции.
Количество решённых задач говорит о надёжности метода с заданными параметрами на решаемом классе тестовых задач.
Чтобы сделать величины, характеризующие надёжность и скорость сходимости, при подсчёте среднего количества испытаний учитывались только решённые задачи.
\end{Russian}

@@

\begin{table}
\begin{center}
\caption{Averaged number of trials executed by sequential methods for solving the test
optimization problems}
  \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
    \hline
    & AGS & AGS\(l\) & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
  \hline
  \(F_{GR}\)     & 193.1 &  \textbf{158.3} & 400.3 & 182.3 & 214.9 & 947.2 & 691.2 & 1257.3 & 1336.8 \\
  \hline
  GKLS 2d Simple &  254.9 & 217.6 & 510.6 & \textbf{189.0} & 255.2 & 556.8 & 356.3 & 952.2 & 1251.5 \\
  \hline
  GKLS 2d Hard   &  728.7 & \textbf{488.0} & 844.7 & 985.4 & 1126.7 & 1042.5 & 1637.9 & 1041.1 & 2532.2 \\
  \hline
  GKLS 3d Simple &  1372.1 & 1195.3 & 4145.8 & \textbf{973.6} & 1477.8 & 4609.2 & 2706.5 & 5956.94 & 3856.1 \\
  \hline
  GKLS 3d Hard   &  3636.1 & \textbf{1930.5} & 6787.0 & 2298.7 & 3553.3 & 5640.1 & 4708.4 & 6914.3 & 7843.2 \\
  \hline
  GKLS 4d Simple &  26654.1 & 11095.7 & 37436.8 & \textbf{7824.3} & 15994.1 & 41514.3 & 21417.9 & 19157.7 & 59895.4 \\
  \hline
  GKLS 4d Hard   &  54536.8 &  \textbf{23167.8} &  73779.3 &  23204.4 &  54489.9 &  80247.2 &  68815.5 &  27466.1 &  109328.1  \\
  \hline
  GKLS 5d Simple &  29810.0 & 11529.0 & 143575.0 & \textbf{7166.5} & 13970.5 & 52647.6 & 34255.3 & 73074.5 & 91580.4 \\
  \hline
  GKLS 5d Hard   &  113129.1 & 67652.7 & 165192.8 & \textbf{66327.4} & 164390.6 & 138766.2 & 116973.1 & 105496.9 & 155123.8 \\
  \hline
  \end{tabular}
  \label{tab:trials}
\end{center}
\end{table}

@@

Сравнение двух параллельных алгоритмов, реализованных в системах Globalizer и MIDACO
было проведено на четырёхмерных задачах GKLS, т.к. задачи большей размерности не поддерживаются
бесплатной версией MIDACO. Как и в случае с последовательными алгоритмами, для кадого класса задач
были изменены параметры методов. Для MIDACO параметр \(focus = -1\). В случае с Globalizer параметр
\(r\) был выбран следуя таблице \ref{tab:params}. С удвоением количества задействованных развёрток
этот параметр уменьшался на 0.3, поскольку увеличение количества развёрток увеличивает и надёжность метода из секции \ref{sub:parallel_evolvents}.

>>>>

В данной работе рассмотрены open-source реализации девяти различных методов глобальной оптимизации.
Все алгоритмы были протестированы на наборе из 900 существенно многоэкстремальных функций, который был сгенерирован с
помощью специализированных генераторов задач \cite{Gaviano2003, grishaginClass}. Помимо сравнения последовательных алгоритмов, на подмножестве из 200 тестовых задач произведено сравнение солвера MIDACO \cite{Schlueter2012} и программной системы Globalizer \cite{globalizerSystem,Strongin2018} в условиях работы на суперкомпьютере Лобачевский.
\end{Russian}

@@

Using the multiple mapping allows solving initial problem (\ref{eq:task}) by parallel solving the
problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1\leqslant s\leqslant S
\]
on a set of intervals $[0,1]$ by the index method. Each one-dimensional problem is solved on a
separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by
particular processor are interpreted as the results of the trials in the rest problems (in the
corresponding points \(x^{k_1},\dots,x^{k_S})\). In this approach, a trial at the point \(x^k \in
[0,1]\) executed in the framework of the \(s\)-th problem, consists in the following sequence of
operations:
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point \( y^k\) (the
blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1\leqslant s\leqslant S\), of the point \(y^k\) and interpret the
trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points
\(x{}^{k_1} ,\dots,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par
The decision rules for the mentioned parallel algorithm, in general, are the same as the rules of the
sequential algorithm (except the method of the trial execution). Each processor has its own copy
of the software realizing the computations of the problem functions and the decision rule of the
index algorithm. For the organization of the interactions among the processors, the queues are
created on each processor, where the processors store the information on the executed iterations
in the form of the tuples: the processor number \(s\), the trial point \(x{}^{k_s}\).
The modifications of the method taking into account multiply criteria are given in \cite{GERGEL2017,Gergel2018}.
\par
The mentioned parallelization scheme was proposed in \cite{Gergel2009} and implemented in the Globalizer system with the use of MPI technology. Main
features of implementation consist in the following:
\begin{itemize}
  \item A separate MPI-process is created for each
 of \(S\) one-dimensional problems being solved, usually, one process per one processor
 employed.
 \item Each process can use $p$ threads to parallel execute $p$ trials, usually one thread per an accessible core.
\end{itemize}

@@

\begin{table}
\begin{center}
\caption{Trials limit for the each of test problems classes}
  \begin{tabular}{|l|{c}|}
    \hline
  Problems class & Trials limit\\
  \hline
  \(F_{GR}\) & 5000 \\
  \hline
  GKLS 2d Simple & 8000 \\
  \hline
  GKLS 2d Hard & 9000 \\
  \hline
  GKLS 3d Simple & 15000 \\
  \hline
  GKLS 3d Hard & 25000 \\
  \hline
  GKLS 4d Simple & 150000 \\
  \hline
  GKLS 4d Hard & 250000 \\
  \hline
  GKLS 5d Simple & 350000 \\
  \hline
  GKLS 5d Hard & 600000 \\
  \hline
  \end{tabular}
  \label{tab:limits}
\end{center}
\end{table}

В качестве характеристик метода оптимизации на каждом из классов будем рассматривать среднее число
испытаний, затраченное для решения одной задачи, и количество решённых задач. Чем меньше число испытаний, тем быстрее метод сходится
к решению, а значит и меньше обращается к потенциально трудоёмкой процедуре вычислений целевой функции.
Количество решённых задач говорит о надёжности метода с заданными параметрами на решаемом классе тестовых задач.
Чтобы сделать величины, характеризующие надёжность и скорость сходимости, независимыми при подсчёте среднего количества испытаний учитывались только решённые задачи.
\end{Russian}

@@

\begin{table}
\begin{center}
\caption{Averaged number of trials executed by sequential methods for solving the test
optimization problems}
\resizebox{\textwidth}{!}{%
  \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
    \hline
    & AGS & AGS\(l\) & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
  \hline
  \(F_{GR}\)     & 193.1 &  \textbf{158.3} & 400.3 & 182.3 & 214.9 & 947.2 & 691.2 & 1257.3 & 1336.8 \\
  \hline
  GKLS 2d Simple &  254.9 & 217.6 & 510.6 & \textbf{189.0} & 255.2 & 556.8 & 356.3 & 952.2 & 1251.5 \\
  \hline
  GKLS 2d Hard   &  728.7 & \textbf{488.0} & 844.7 & 985.4 & 1126.7 & 1042.5 & 1637.9 & 1041.1 & 2532.2 \\
  \hline
  GKLS 3d Simple &  1372.1 & 1195.3 & 4145.8 & \textbf{973.6} & 1477.8 & 4609.2 & 2706.5 & 5956.94 & 3856.1 \\
  \hline
  GKLS 3d Hard   &  3636.1 & \textbf{1930.5} & 6787.0 & 2298.7 & 3553.3 & 5640.1 & 4708.4 & 6914.3 & 7843.2 \\
  \hline
  GKLS 4d Simple &  26654.1 & 11095.7 & 37436.8 & \textbf{7824.3} & 15994.1 & 41514.3 & 21417.9 & 19157.7 & 59895.4 \\
  \hline
  GKLS 4d Hard   &  54536.8 &  \textbf{23167.8} &  73779.3 &  23204.4 &  54489.9 &  80247.2 &  68815.5 &  27466.1 &  109328.1  \\
  \hline
  GKLS 5d Simple &  29810.0 & 11529.0 & 143575.0 & \textbf{7166.5} & 13970.5 & 52647.6 & 34255.3 & 73074.5 & 91580.4 \\
  \hline
  GKLS 5d Hard   &  113129.1 & 67652.7 & 165192.8 & \textbf{66327.4} & 164390.6 & 138766.2 & 116973.1 & 105496.9 & 155123.8 \\
  \hline
\end{tabular}}
  \label{tab:trials}
\end{center}
\end{table}

@@

Сравнение двух параллельных алгоритмов, реализованных в системах Globalizer и MIDACO
было проведено на четырёхмерных задачах GKLS, т.к. задачи большей размерности не поддерживаются
бесплатной версией MIDACO. Как и в случае с последовательными алгоритмами, для кадого класса задач
были изменены параметры методов. Для MIDACO параметр \(focus = -1\). В случае с Globalizer параметр
\(r\) был выбран следуя таблице \ref{tab:params}. С удвоением количества задействованных развёрток
этот параметр уменьшался на 0.3, поскольку увеличение количества развёрток увеличивает и надёжность метода из секции \ref{sub:parallel_evolvents}
(подробнее см. \cite{SOVRASOV2018}).
